Modelling the emergence of whisker barrels

### 2019-12-13 paper/briefpaper_stuedits/brief.tex
paper draft updates

\dropcap{S}patial patterns in neural connectivity provide clues about the constraints under which brains evolve and develop \citep{purves_iterated_1992}. Perhaps the most distinctive pattern of neural connectivity can be found in the rodent barrel cortex \cite{woolsey_structural_1970}. The barrels are large cortical columns, identifiable from birth in layer 4 of primary somatosensory cortex as dense clusters of thalamocortical axons, which are enclosed by borders a few neurons thick from postnatal day 3 \citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the barrels constitute a somatotopic map of the whiskers, with cells of adjacent barrels responding most strongly and quickly to deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel patterning reflects subcortical whisker maps comprising cell aggregates called barrelettes in the brainstem and barreloids in the thalamus \cite{van_der_loos_barreloids_1976}.

Barrel patterning requires afferent input from whisker stimulation and thalamic calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex network of axon guidance molecules such as ephrin-A5 and A7 and adhesion molecules such as cadherin-6 and 8 \citep{miller_epha7-ephrin-a5_2006}.
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17 and transcription factors Emx2, Pax6, Sp8, and Coup-tf1 \citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed in gradients that mark orthogonal axes and can be manipulated to stretch, shrink, shift, and even duplicate barrels \cite{assimacopoulos_fibroblast_2012}.

@@

Alternatively, reaction-diffusion dynamics could generate a Voronoi tessellation without pre-arranged centres, by amplifying characteristic modes in a noisy initial distribution of axon branches, as a net effect of short-range cooperative and longer-range competitive interactions. Accordingly, the barrel pattern would be determined by the relative strength of these interactions and by the shape of the cortical field boundary. However, intrinsic cortical dynamics alone cannot account for the topographic correspondence between domains in the thalamus and cortex, or the irregular sizes and specific arrangement of the barrels in rows and arcs, or the influence of gene expression gradients.

@@

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a reaction-diffusion style model of how extrinsic signalling gradients can constrain the emergence of distinct fields from intrinsic cortical dynamics. Their model defines how the density of connections $c(x,t)$, and axon branches $a(x,t)$, interact at time $t$, along a 1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by $i$. The model was derived from the assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally coupled. Extending the original 1D model to simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and  model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{k=1}^{N} c_{k}\right)[a_i]^l.
\end{equation}
%
Accordingly, when the total density of connections sums to one, the density of thalamocortical connections decays at rate $\alpha$. Otherwise the synaptic connection density increases non-linearly ($l>1$) with the axon branching density. Changes in axon branching are modelled as 
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by $\nabla\vcdot$) of the quantity in parentheses, which is referred to as the `flux of axonal branching'. The flux represents diffusion across the cortical sheet, at rate $D$, and the influence of $M$ molecular signalling fields, $\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which may be positive or negative in order that axons may branch in the direction of either higher or lower  concentrations. The second term on the right represents an assumption that axon branching decreases when synaptogenesis increases, and vice versa, i.e., that axon branches are required for synaptic connections to form. Here $\chi_i=0$ is a placeholder.

@@

First we verified that all results established for a 1D axis could be reproduced using our extension to a 2D cortical sheet. Using an elliptical boundary with $M=3$ guidance gradients aligned to the longer axis, $N=5$ thalamocortical projections gave rise to five distinct cortical fields at locations that preserved the topographic ordering defined by the original $\gamma$ values of \cite{karbowski_model_2004}. However, we found that specifying $N$ ordered areas required $M\approx N$ signalling fields. As the number of distinct guidance fields is unlikely to approach the number of barrels, further extensions were required.

The term in parentheses in Eq.~\ref{eq:dc} represents competition between thalamic projections for a limited availability of cortical connections. To introduce competition also in terms of axon branching we redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branching of other projections is high. Divisive normalization is used to keep the axon branch density bounded on each iteration to the initial density  at $t=0$,

@@

The only differences between thalamic projections are the interaction strengths, $\gamma$, hence any reliable differences between emergent cortical fields must be due to differences in these values only. We chose $\gamma$ values by speculating that interaction strengths with ephrin gradients in the cortical subplate are determined by sampling from similar thalamic gradients at the origin of each thalamic projection, i.e., at each barreloid centre. Two orthogonal linear thalamic gradients were defined, from which 52 pairs of $\gamma$ values were sampled, at the coordinates of 52 barreloid centers recreated from  \cite{haidarliu_size_2001}.

>>>>

\dropcap{S}patial patterns in neural connectivity provide clues about the constraints under which brains evolve and develop \citep{purves_iterated_1992}. Perhaps the most distinctive pattern can be found in the rodent barrel cortex \cite{woolsey_structural_1970}. The barrels are large cortical columns, identifiable from birth in layer 4 of primary somatosensory cortex as dense clusters of thalamocortical axons, which are enclosed by borders a few neurons thick from postnatal day 3 \citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the barrels constitute a somatotopic map of the whiskers, with cells of adjacent barrels responding most strongly and quickly to deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel patterning reflects subcortical whisker maps comprising cell aggregates called barrelettes in the brainstem and barreloids in the thalamus \cite{van_der_loos_barreloids_1976}.

Barrel formation requires afferent input from whisker stimulation and thalamic calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex network of axon guidance molecules such as ephrin-A5 and A7 and adhesion molecules such as cadherin-6 and 8 \citep{miller_epha7-ephrin-a5_2006}.
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17 and transcription factors Emx2, Pax6, Sp8, and Coup-tf1 \citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed in gradients that mark orthogonal axes and can be manipulated to stretch, shrink, shift, and even duplicate barrels \cite{assimacopoulos_fibroblast_2012}.

@@

Alternatively, reaction-diffusion dynamics could generate a Voronoi tessellation without pre-arranged centres, by amplifying characteristic modes in a noisy initial distribution of axon branches, as a net effect of short-range cooperative and longer-range competitive interactions. Accordingly, the barrel pattern would be determined by the relative strength of these interactions and by the shape of the cortical field boundary. However, intrinsic cortical dynamics alone cannot account for the topographic correspondence between thalamic and cortical domains, the irregular sizes and specific arrangement of the barrels in rows and arcs, or the influence of gene expression gradients.

@@

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a reaction-diffusion style model of how extrinsic signalling gradients can constrain the emergence of distinct fields from intrinsic cortical dynamics. Their model defines how the density of connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by $i$. The model was derived from the assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally coupled. Extending the original 1D model to simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and  model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{k=1}^{N} c_{k}\right)[a_i]^l.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one, connections decay at rate $\alpha$. Otherwise the connection density increases non-linearly ($l>1$) with the density of axon branching. Axon branching is modelled as 
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by $\nabla\vcdot$) of the quantity in parentheses, which is referred to as the `flux' of axonal branching. The flux represents diffusion across the cortical sheet, at rate $D$, and the influence of $M$ molecular signalling fields, $\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which may be positive or negative in order that axons may branch in the direction of either higher or lower  concentrations. The second term on the right represents an assumption that axon branching decreases when synaptogenesis increases, and vice versa, i.e., that axon branches are required for synaptic connections to form. Here $\chi_i=0$ is a placeholder.

@@

First we verified that all results established by \cite{karbowski_model_2004} for a 1D axis could be reproduced using our extension to a 2D cortical sheet. Using an elliptical boundary with $M=3$ offset guidance gradients aligned to the longer axis, $N=5$ thalamocortical projections gave rise to five distinct cortical fields at locations that preserved the topographic ordering defined by the original $\gamma$ values. However, we found that specifying $N$ ordered areas required $M\approx N$ signalling fields. As the number of distinct guidance fields is unlikely to approach the number of barrels, further extensions were required.

The term in parentheses in Eq.~\ref{eq:dc} represents competition between thalamocortical projections for a limited availability of cortical connections. To introduce competition also in terms of axon branching we redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other projections are dense. Divisive normalization keeps the axon branch density bounded on each iteration to the initial density  at $t=0$;

@@

The only differences between thalamic projections are their strengths of interaction with the guidance fields, $\gamma$, hence any reliable differences between emergent cortical fields must be due to differences in these values only. We speculate that a projection interacts with a given ephrin field in the cortical subplate with a strength determined by the concentration of a similar molecule at its thalamic origin, i.e., the putative barreloid center. As such, two orthogonal linear thalamic gradients were defined, from which 52 pairs of $\gamma$ values were sampled, at the coordinates of 52 barreloid centers recreated from  \cite{haidarliu_size_2001}.
